---
title: "Mini-Project #1 - Netflix Top 10"
format:
  html:
    css: styles.css
    code-fold: false
    code-tools: true
    code-summary: "Show code"
execute:
  warning: true
  message: true
---

## Objective

For this assignment, we are given the task of analyzing Netflix's Global Top 10 and Country Top 10 datasets. Below we will acquire, clean, and explore the data, then answer exploratory questions and draft press releases.

## Acquire Data

```{r}
if (!dir.exists(file.path("data", "mp01"))) {
  dir.create(file.path("data", "mp01"), showWarnings = FALSE, recursive = TRUE)
}

GLOBAL_TOP_10_FILENAME  <- file.path("data", "mp01", "global_top10_alltime.csv")
COUNTRY_TOP_10_FILENAME <- file.path("data", "mp01", "country_top10_alltime.csv")

if (!file.exists(GLOBAL_TOP_10_FILENAME)) {
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv",
                destfile = GLOBAL_TOP_10_FILENAME)
}
if (!file.exists(COUNTRY_TOP_10_FILENAME)) {
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv",
                destfile = COUNTRY_TOP_10_FILENAME)
}
```

## Data Import
```{r}
if (!require("tidyverse")) install.packages("tidyverse")
library(readr)
library(dplyr)
library(stringr)
library(DT)
```

```{r}
GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME)
COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME)
```

## Data Cleaning/Reformatting
```{r}
GLOBAL_TOP_10 <- GLOBAL_TOP_10 |>
mutate(season_title = if_else(season_title == "N/A", NA_character_, season_title))
```

```{r}
COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME,
                          na = "N/A")
```

## Intial Data Exploration
```{r}
library(DT)
GLOBAL_TOP_10 |> 
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE))
```

```{r}
library(stringr)
format_titles <- function(df){
    colnames(df) <- str_replace_all(colnames(df), "_", " ") |> str_to_title()
    df
}

GLOBAL_TOP_10 |> 
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

```{r}
GLOBAL_TOP_10 |> 
    select(-season_title) |>
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

```{r}
GLOBAL_TOP_10 |> 
    mutate(`runtime_(minutes)` = round(60 * runtime)) |>
    select(-season_title, 
           -runtime) |>
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

## Exploratory Questions
1. How many different countries does Netflix operate in? 
```{r}
countries <- COUNTRY_TOP_10 |> 
  summarise(num_countries = n_distinct(country_name))
```
Netflix operates in **`r countries`** countries.

2. Which non-English-language film has spent the most cumulative weeks in the global top 10? How many weeks did it spend?
```{r}
nonEnglish <- GLOBAL_TOP_10 |> 
  filter(category == "Films (Non-English)") |>
  arrange(desc(cumulative_weeks_in_top_10)) |>
  slice(1) |>
  select(show_title, cumulative_weeks_in_top_10)
```
The non-English film with the most cumulative weeks in the global Top 10 is **`r nonEnglish$show_title`**, 
with **`r nonEnglish$cumulative_weeks_in_top_10`** weeks.

3. What is the longest film (English or non-English) to have ever appeared in the Netflix global Top 10? How long is it in minutes?
```{r}
longest <- GLOBAL_TOP_10 |>
  filter(str_detect(category, "Films")) |>
  mutate(runtime_minutes = runtime * 60) |>
  arrange(desc(runtime_minutes)) |> 
  slice(1) |>
  select(show_title, runtime_minutes)
```
The longest film to have ever appeared in the Netflix global Top 10 was **`r longest$show_title`** with **`r round(longest$runtime_minutes)`** minutes.

4. For each of the four categories, what program has the most total hours of global viewership?
```{r}
top_by_category <- GLOBAL_TOP_10 |>
  group_by(category, show_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE), .groups = "drop_last") |>
  slice_max(order_by = total_hours, n = 1, with_ties = FALSE) |>
  ungroup()

```
The programs with the most total hours of global viewership are **`r top_by_category$show_title[top_by_category$category == "Films (English)"]`** for English films, **`r top_by_category$show_title[top_by_category$category == "Films (Non-English)"]`** for non-English films, **`r top_by_category$show_title[top_by_category$category == "TV (English)"]`** for English TV, and **`r top_by_category$show_title[top_by_category$category == "TV (Non-English)"]`** for non-English TV.

5. Which TV show had the longest run in a country’s Top 10? How long was this run and in what country did it occur?
```{r}
longest_tv_run <- COUNTRY_TOP_10 |>
  filter(category == "TV") |>
  group_by(country_name, show_title) |>
  summarise(longest_run = max(cumulative_weeks_in_top_10, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(longest_run)) |>
  slice(1)   # Keep only the very top entry
```
The TV show that had the longest run in a country’s Top 10 is **`r longest_tv_run$show_title`**, which appeared in **`r longest_tv_run$country_name`**  for **`r longest_tv_run$longest_run`** weeks.

6. Netflix provides over 200 weeks of service history for all but one country in our data set. Which country is this and when did Netflix cease operations in that country?
```{r}
dropped_country <- COUNTRY_TOP_10 |>
  group_by(country_name) |>
  summarise(max_week = max(week, na.rm = TRUE)) |>
  arrange(max_week) |>
  slice(1)
```
The country where Netflix ceased operations the earliest is **`r dropped_country$country_name`**, in the week of **`r dropped_country$max_week`**.

7. What is the total viewership of the TV show Squid Game? Note that there are three seasons total and we are looking for the total number of hours watched across all seasons.
```{r}
squid_game_viewership <- GLOBAL_TOP_10 |>
  filter(str_detect(show_title, "Squid Game")) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE))
```
The total viewership of the TV show *Squid Game* across all seasons is **`r scales::comma(squid_game_viewership$total_hours)`** hours.

8. The movie Red Notice has a runtime of 1 hour and 58 minutes. Approximately how many views did it receive in 2021?
```{r}
red_notice_views_2021 <- GLOBAL_TOP_10 |>
  filter(show_title == "Red Notice", year(week) == 2021) |>
  summarise(total_views = sum(weekly_hours_viewed, na.rm = TRUE) / ((1 + 58 / 60))) |>
  mutate(total_views = round(total_views))
```
The movie *Red Notice* received approximately **`r scales::comma(red_notice_views_2021$total_views)`** views in 2021.

9. How many Films reached Number 1 in the US but did not originally debut there? 
```{r}
library(dplyr)
library(stringr)

us_films <- COUNTRY_TOP_10 |>
  filter(country_name == "United States",
         str_starts(category, "Films")) |>
  group_by(show_title) |>
  summarise(
    debut_week = min(week, na.rm = TRUE),
    debut_rank = weekly_rank[which.min(week)],
    hit_one = any(weekly_rank == 1, na.rm = TRUE),
    last_week_one = if (any(weekly_rank == 1, na.rm = TRUE))
      max(week[weekly_rank == 1], na.rm = TRUE)
    else as.Date(NA),
    .groups = "drop"
  )

climbers <- us_films |>
  filter(hit_one, debut_rank > 1) |>
  arrange(desc(last_week_one))

films_count <- nrow(climbers)
latest_row  <- slice_head(climbers, n = 1)
latest_film <- latest_row$show_title
latest_when <- latest_row$last_week_one

```
A total of **`r films_count`** films reached Number 1 in the United States **after** debuting below the top spot.  
The most recent film to do this was **`r latest_film`**, which reached **Number 1** on **`r latest_when`**.

10. Which TV show/season hit the top 10 in the most countries in its debut week? In how many countries did it chart?
```{r}
library(dplyr)
library(stringr)

tv_debut_spread <- COUNTRY_TOP_10 |>
  filter(str_starts(category, "TV")) |>
  group_by(show_title, season_title) |>
  mutate(debut_week = min(week, na.rm = TRUE)) |>
  ungroup() |>
  filter(week == debut_week) |>
  group_by(show_title, season_title) |>
  summarise(
    countries = n_distinct(country_name),
    debut_week = first(debut_week),
    .groups = "drop"
  ) |>
  slice_max(countries, n = 1, with_ties = FALSE)

top_title     <- tv_debut_spread$show_title
top_season    <- tv_debut_spread$season_title
top_countries <- tv_debut_spread$countries
top_week      <- tv_debut_spread$debut_week
```
The TV show/season that hit the Top 10 in the most countries in its debut week was **`r top_title` — `r top_season`**, charting in **`r top_countries`** countries during the week of **`r top_week`**.

## Press Releases
### Stranger Things Season 5: The Global Phenomenon Returns
Netflix proudly announces the upcoming fifth and final season of Stranger Things, one of the most-watched English-language TV shows in the platform’s history. Across its previous four seasons, the series has amassed over 2.9 billion total viewing hours, proving its unmatched global appeal. Stranger Things held a spot in the Top 10 for more than 350 weeks worldwide, appearing in over 90 countries. While other major hits like Wednesday and The Witcher have seen massive success, Stranger Things continues to lead as the gold standard for global engagement and storytelling. Fans everywhere are eagerly awaiting the final showdown in Hawkins later this year.

### Netflix Celebrates Blockbuster Growth in India
India’s love for local storytelling continues to fuel Netflix’s growth in the region. Hindi-language films and TV shows have now spent over 200 weeks in India’s Top 10 charts, with recent hits like The Secret of Devkaali, achieving record-breaking viewership. Total hours viewed in India have surpassed 180 million, showing an expanding subscriber base and rising engagement across the country. With India now recognized as one of Netflix’s fastest-growing markets, the company continues to invest in local productions that connect deeply with audiences at home and abroad.

### Japan’s Thrillers Take Center Stage on Netflix
Netflix Japan is riding a wave of suspense-driven storytelling, as homegrown thrillers climb the global charts. Titles such as Exit 8 have captivated millions with it's blend of psychological and cinematic tension. In 2025, Japanese thrillers appeared in the Top 10 across more than 70 countries, marking a record year for international engagement. Viewers are drawn to Japan’s unique ability to fuse quiet drama with gripping mystery, creating shows that stay in the global Top 10 for weeks at a time. As Netflix invests further in Japan’s thriller genre, these stories continue to prove that intrigue knows no language barrier.
